---
title: "ESM 244 Lab 1"
author: "Claire Meuter"
date: "2023-01-13"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE, warning = FALSE, message = FALSE}
#include = true means we can see our setup chunk and what packages were installed
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
#command shift return to run the whole chunk at once 
```


### Read in our data 
```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
#read.csv is the built in, read_csv is an improved version in the tidyverse, it adds additional functionality. read_csv will generally eb faster anyways 
```


#exploratory in console
#can use summary(sf_trees) to get an idea of what were seeing in the data set 
#can also do names(sf_trees)
#can also do sf_trees %>% group_by(legal_status) %>% summarize(tree_count = n())

#more data exploration
```{r}
top_5_status <- sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n()) %>%
  slice_max(tree_count, n = 5) %>% #slice_max function finds the variables that are the maximum and only keep the top 5 (in this case bc n =5)
  arrange(-tree_count)

#make a plot of our data 
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y =tree_count)) +
  geom_col() +
  labs(y = 'Tree Count', x ='Legal Status') + 
  coord_flip() + #flip x and y axis 
  theme_minimal() #adding a theme to make it look better
```
### Keep observation where legal status is "permited site" and caretaker is "MTA" 
```{r}
# sf_trees$legal_status %>% unique()   to see list of legal_status options 
permitted_mta <- sf_trees %>% 
  filter(legal_status == "Permitted Site" & caretaker == "MTA") 
```

### Keep Blackwood acacia trees, only legal_status, date, latitude longitude columns 
```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude) #we don't need to include species anymore, so only keep the columns relevant to our map we'll make 

### make a plot of SF acacia 
ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) + geom_point()


```
### Use tidyr::seperate() to seperate one column into multiple, and tidy::unite() to rejoin
```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')

#reversing that: putting two columns together 

sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id, legal_status, species, sep = "_NEW_")

```

